---
title: "STAT 1361 Final Project Code"
author: "Elizabeth Kwee"
date: "2026-04-21"
output:
  pdf_document: default
  html_document: default
---
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(regclass) #VIF
library(leaps) #subset selection
library(caret) #cross validation

library(glmnet) #ridge and lasso regression
library(tree) #decision tree
library(randomForest) #random forest
```


```{r}
#Read in data
train = read.csv("df_train.csv", header = T)
test = read.csv("df_test.csv", header = T)

#View data
head(train, 5)
```

##Exploratory Data Analysis
```{r}
#Convert categorical variables to numeric
train.numeric = train
train.numeric$neighbourhood_cleansed = as.numeric(as.factor(train$neighbourhood_cleansed))
train.numeric$room_type = as.numeric(as.factor(train$room_type))
```

#Univariate Analysis
```{r, message=FALSE, warning=FALSE}
#Price (target variable)
ggplot(train, aes(x = price)) +
  geom_boxplot() +
  labs(title = "Price boxplot with outliers", x = "Price")

ggplot(train, aes(x = price)) +
  geom_boxplot(outliers = F) +
  labs(title = "Price boxplot without outliers", x = "Price")

#Neighbourhood_cleansed
ggplot(train, aes(x=neighbourhood_cleansed)) +
  geom_bar(fill="lightblue") +
  coord_flip() +
  labs(title = "Bar chart of neighborhoods", x = "Neighborhoods")

#Room type
ggplot(train, aes(x=room_type)) +
  geom_bar(fill="lightblue") +
  labs(title = "Bar chart of room types", x = "Room types")

#Bathrooms
ggplot(train, aes(x = bathrooms)) +
  geom_histogram(bins = 8, fill="lightblue", col="black") +
  labs(title = "Bathrooms histogram", x = "Number of bathrooms")

#Bedrooms
ggplot(train, aes(x = bedrooms)) +
  geom_histogram(bins = 15, fill="lightblue", col="black") +
  labs(title = "Bedrooms histogram", x = "Number of bedrooms")

#Accommodates
ggplot(train, aes(x = accommodates)) +
  geom_histogram(bins = 15, fill="lightblue", col="black") +
  labs(title = "Accommodates histogram", x = "Number of people accommadatable")

#Minimum_nights
ggplot(train, aes(x = minimum_nights)) +
  geom_boxplot() +
  labs(title = "Min nights boxplot", x = "Min nights")

ggplot(train, aes(x = minimum_nights)) +
  geom_histogram(fill="lightblue", col="black") +
  labs(title = "Min nights histogram", x = "Min nights")

#Availability_365
ggplot(train, aes(x = availability_365)) +
  geom_histogram(bins = 50, fill="lightblue", col="black") +
  labs(title = "Availability in a year histogram", x = "Days available")

#Number_of_reviews
ggplot(train, aes(x = number_of_reviews)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Number of reviews histogram", x = "Number of reviews")

#Review_scores_rating
ggplot(train, aes(x = review_scores_rating)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Review scores histogram", x = "Scores")

#Review_scores_cleanliness
ggplot(train, aes(x = review_scores_cleanliness)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Review scores cleanliness histogram", x = "Cleanliness")

#Review_scores_location
ggplot(train, aes(x = review_scores_location)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Review scores location histogram", x = "Location")

#Host_is_superhost
ggplot(train, aes(x = as.factor(host_is_superhost))) +
  geom_bar(fill = "lightblue") +
  labs(title ="Host is superhost", x = "Superhost (1) or not (0)")

#Instant_bookable
ggplot(train, aes(x = as.factor(instant_bookable))) +
  geom_bar(fill = "lightblue") +
  labs(title ="Instant bookable", x = "Instant (1) or not (0)")

#Host_identity_verified
ggplot(train, aes(x = as.factor(host_identity_verified))) +
  geom_bar(fill = "lightblue") +
  labs(title ="Host identity verified", x = "Verified (1) or not (0)")

#Calculated_host_listings_count
ggplot(train, aes(x = calculated_host_listings_count)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Number of host listings histogram", x = "Number of listings")

#Dist_to_downtown
ggplot(train, aes(x = dist_to_downtown)) +
  geom_histogram(fill = "lightblue", col = "black") +
  labs(title ="Distance to downtown histogram", x = "Distance (mi)")

```

```{r}
summary(train$price)
outlier_values = boxplot.stats(train$price)$out
length(outlier_values)
```

##Bivariate analysis
#Price vs. variables
```{r}
for (i in 1:16){
  plot(train.numeric[,i], train.numeric$price, xlab = colnames(train.numeric)[i], ylab="Price")
}
```


#Multivariate Analysis
```{r}
#Correlation between pairs of variables
cor(train.numeric)

#Check for potential multicollinearity (none clearly observed)
full_lm.numeric = lm(price ~ ., data = train.numeric)
print("VIF Values")
VIF(full_lm.numeric)
```

```{r}
#Influential points
plot(full_lm.numeric, which=4)
abline(h = 4/nrow(train.numeric), col="red")

cd = cooks.distance(full_lm.numeric)
influential.points = which(cd > 4/nrow(train.numeric))
length(influential.points)
```

```{r}
#Analyzing influential points
sort(cd, decreasing = T)[1:4]
train[635,]
train[2139,]
train[1596,]
train[334,]
```

```{r}
train.delete = train[-c(635, 2139, 1596, 334),]
train.delete.numeric = train.delete
train.delete.numeric$neighbourhood_cleansed = as.numeric(as.factor(train.delete.numeric$neighbourhood_cleansed))
train.delete.numeric$room_type = as.numeric(as.factor(train.delete.numeric$room_type))
```


\newpage
##Model analysis with all predictors
#Full linear model (base instance)
```{r}
full_lm = lm(price ~ ., data = train)
full_lm.ypred = predict(full_lm, train)
full_lm.mse = mean((full_lm.ypred - train$price)^2)
sqrt(full_lm.mse)
summary(full_lm)
```

#Full linear model with influential points removed
```{r}
full_lm.rem = lm(price ~ ., data = train.delete)
full_lm.rem.ypred = predict(full_lm.rem, train.delete)
full_lm.rem.mse = mean((full_lm.rem.ypred - train.delete$price)^2)
sqrt(full_lm.rem.mse)
summary(full_lm.rem)
```

```{r}
par(mfrow=c(2,2))
plot(full_lm.rem)
```

#GLM with log link
```{r}
#Full dataset
log_glm = glm(price ~ ., data = train, family = gaussian(link="log"))
log_glm.ypred <- predict(log_glm, train, type = "response")
log_glm.mse = mean((log_glm.ypred - train$price)^2)
paste0("GLM with log link RMSE: ", round(sqrt(log_glm.mse),2))

#Deleted 4 influential points
log_glmm4 = glm(price ~ ., data = train.delete, family = gaussian(link="log"))
log_glm.ypredm4 <- predict(log_glm, train.delete, type = "response")
log_glm.msem4 = mean((log_glm.ypredm4 - train.delete$price)^2)
paste0("GLM with log link RMSE (-4): ", round(sqrt(log_glm.msem4),2))
```

```{r}
par(mfrow=c(2,2))
plot(log_glm)
summary(log_glm)
```


#Ridge and lasso regression (all predictors) w/CV
```{r}
set.seed(1)
x = model.matrix(price ~ ., data = train)[,-1] 
y = train$price

#all data points
ridge.cv = cv.glmnet(x,y,alpha=0)
best.ridge.lambda = ridge.cv$lambda.min
ridge_model = glmnet(x, y, alpha=0, lambda = best.ridge.lambda)
ridge.ypred = predict(ridge_model, newx = x, lambda = best.ridge.lambda)
ridge.mse = mean((ridge.ypred - y)^2)
paste0("Full ridge RSME: ", round(sqrt(ridge.mse), 2))

lasso.cv = cv.glmnet(x,y,alpha=1)
best.lasso.lambda = lasso.cv$lambda.min
lasso_model = glmnet(x, y, alpha=1, lambda = best.lasso.lambda)
lasso.ypred = predict(lasso_model, newx = x, lambda = best.lasso.lambda)
lasso.mse = mean((lasso.ypred - y)^2)
paste0("Full lasso RSME: ", round(sqrt(lasso.mse), 2))


#deleted 4 influential points
xm4 = model.matrix(price ~ ., data = train)[-c(635, 2139, 1596, 334),-1]
ym4 = y[-c(635, 2139, 1596, 334)]

ridge.cvm4 = cv.glmnet(xm4,ym4,alpha=0)
best.ridge.lambdam4 = ridge.cvm4$lambda.min
ridge_modelm4 = glmnet(xm4, ym4, alpha=0, lambda = best.ridge.lambdam4)
ridge.ypredm4 = predict(ridge_modelm4, newx = xm4, lambda = best.ridge.lambdam4)
ridge.msem4 = mean((ridge.ypredm4 - ym4)^2)
paste0("Full ridge RSME (-4): ", round(sqrt(ridge.msem4), 2))

lasso.cvm4 = cv.glmnet(xm4,ym4,alpha=1)
best.lasso.lambdam4 = lasso.cvm4$lambda.min
lasso_modelm4 = glmnet(xm4, ym4, alpha=1, lambda = best.lasso.lambdam4)
lasso.ypredm4 = predict(lasso_modelm4, newx = xm4, lambda = best.lasso.lambdam4)
lasso.msem4 = mean((lasso.ypredm4 - ym4)^2)
paste0("Full lasso RSME (-4): ", round(sqrt(lasso.msem4), 2))

```


#Decision tree
```{r}
set.seed(1)
#Full dataset
tree_model = tree(price ~ ., data = train.numeric)
tree_model.ypred = predict(tree_model, train.numeric)
tree.mse = mean((tree_model.ypred - train.numeric$price)^2)
paste0("Full tree RSME: ", round(sqrt(tree.mse), 2))

plot(tree_model)
text(tree_model, cex=0.8)

#Deleted 4 influential points
tree_modelm4 = tree(price ~ ., data = train.delete.numeric)
tree_model.ypredm4 = predict(tree_modelm4, train.delete.numeric)
tree.msem4 = mean((tree_model.ypredm4 - train.delete.numeric$price)^2)
paste0("Full tree RSME (-4): ", round(sqrt(tree.msem4), 2))

plot(tree_modelm4)
text(tree_modelm4, cex=0.8)

```


#Pruned tree with CV
```{r}
cv.airbnb = cv.tree(tree_model)
plot(cv.airbnb$size, cv.airbnb$dev, type = "b")
which.min(cv.airbnb$dev)
pruned.tree_model = prune.tree(tree_model, best = 6)
plot(pruned.tree_model)
text(pruned.tree_model)

pruned.tree.ypred = predict(pruned.tree_model, train.numeric)
pruned.mse = mean((pruned.tree.ypred - train.numeric$price)^2)
paste0("Pruned tree RSME: ", round(sqrt(pruned.mse), 2))
```


#Random forest
```{r}
set.seed(1)

#Full dataset
rf_model = randomForest(price ~ ., data = train.numeric, importance=TRUE)
rf.ypred = predict(rf_model, train.numeric)
rf.mse = mean((rf.ypred - train.numeric$price)^2)
paste0("Random Forest RMSE: ", round(sqrt(rf.mse),2))

print(rf_model)
varImpPlot(rf_model)
importance(rf_model)


#Deleted observations
rf_modelm4 = randomForest(price ~ ., data = train.delete.numeric, importance=TRUE)
rf.ypredm4 = predict(rf_modelm4, train.delete.numeric)
rf.msem4 = mean((rf.ypredm4 - train.delete.numeric$price)^2)
paste0("Random Forest RMSE (-4): ", round(sqrt(rf.msem4),2))

print(rf_modelm4)
varImpPlot(rf_modelm4)
importance(rf_modelm4)
```


\newpage
##Model analysis with subset of predictors
#Subset selection to obtain most significant variables
```{r}
#Full dataset
regfit.forward = regsubsets(price ~ ., data = train, nvmax = 32, method = "forward")
reg.summary = summary(regfit.forward)
reg.summary
plot(reg.summary$rss , xlab = "Number of Variables", ylab = "RSS", type = "l")
plot(reg.summary$adjr2 , xlab = "Number of Variables", ylab = " Adjusted RSq", type = "l")
which.max(reg.summary$adjr2) #adj r2 maximized with 17 predictors
coef(regfit.forward, 17)

#Deleted 4 influential points
regfit.forwardm4 = regsubsets(price ~ ., data = train.delete, nvmax = 32, method = "forward")
reg.summarym4 = summary(regfit.forwardm4)
plot(reg.summarym4$rss , xlab = "Number of Variables", ylab = "RSS", type = "l")
plot(reg.summarym4$adjr2 , xlab = "Number of Variables", ylab = " Adjusted RSq", type = "l")
which.max(reg.summarym4$adjr2) #adj r2 maximized with 23 predictors
coef(regfit.forwardm4, 23)
```


#Linear model (subset)
```{r}
#Full dataset
part_lm = lm(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train)
part_lm.ypred = predict(part_lm, train)
part_lm.mse = mean((part_lm.ypred - train$price)^2)
paste0("Partial linear model RMSE: ", round(sqrt(part_lm.mse),2))

#Deleted 4 points
part_lmm4 = lm(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train.delete)
part_lm.ypredm4 = predict(part_lmm4, train.delete)
part_lm.msem4 = mean((part_lm.ypredm4 - train.delete$price)^2)
paste0("Partial linear model RMSE (-4): ", round(sqrt(part_lm.msem4),2))
```


#GLM with log link (subset)
```{r}
log_part_glm = glm(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train, family = gaussian(link="log"))
log_part_glm.ypred <- predict(log_part_glm, train, type = "response")
log_part_glm.mse = mean((log_part_glm.ypred - train$price)^2)
paste0("Partial GLM with log link RMSE: ", round(sqrt(log_part_glm.mse),2))

#Deleted 4 influential points
log_part_glmm4 = glm(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train.delete, family = gaussian(link="log"))
log_part_glm.ypredm4 <- predict(log_part_glm, train.delete, type = "response")
log_part_glm.msem4 = mean((log_part_glm.ypredm4 - train.delete$price)^2)
paste0("Partial GLM with log link RMSE (-4): ", round(sqrt(log_part_glm.msem4),2))
```


#Ridge and lasso regression (subset) w/CV
```{r}
set.seed(1)
#Full dataset
x2 = model.matrix(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train)[,-1] 

part_ridge.cv = cv.glmnet(x2,y,alpha=0)
part_best.ridge.lambda = part_ridge.cv$lambda.min
part_ridge_model = glmnet(x2, y, alpha=0, lambda = part_best.ridge.lambda)
part_ridge.ypred = predict(part_ridge_model, newx = x2, lambda = part_best.ridge.lambda)
part_ridge.mse = mean((part_ridge.ypred - y)^2)
paste0("Partial ridge RSME: ", round(sqrt(part_ridge.mse), 2))

part_lasso.cv = cv.glmnet(x2,y,alpha=1)
part_best.lasso.lambda = part_lasso.cv$lambda.min
part_lasso_model = glmnet(x2, y, alpha=1, lambda = part_best.lasso.lambda)
part_lasso.ypred = predict(part_lasso_model, newx = x2, lambda = part_best.lasso.lambda)
part_lasso.mse = mean((part_lasso.ypred - y)^2)
paste0("Partial lasso RSME: ", round(sqrt(part_lasso.mse), 2))

#4 influential points removed
x2m4 = model.matrix(price ~ neighbourhood_cleansed + room_type + bathrooms + bedrooms + number_of_reviews + accommodates + availability_365 + review_scores_cleanliness + review_scores_location + host_is_superhost + instant_bookable + host_identity_verified + dist_to_downtown, data = train.delete)[,-1] 

part_ridge.cvm4 = cv.glmnet(x2m4,ym4,alpha=0)
part_best.ridge.lambdam4 = part_ridge.cvm4$lambda.min
part_ridge_modelm4 = glmnet(x2m4, ym4, alpha=0, lambda = part_best.ridge.lambdam4)
part_ridge.ypredm4 = predict(part_ridge_modelm4, newx = x2m4, lambda = part_best.ridge.lambdam4)
part_ridge.msem4 = mean((part_ridge.ypredm4 - ym4)^2)
paste0("Partial ridge RSME (-4): ", round(sqrt(part_ridge.msem4), 2))

part_lasso.cvm4 = cv.glmnet(x2m4,ym4,alpha=1)
part_best.lasso.lambdam4 = part_lasso.cvm4$lambda.min
part_lasso_modelm4 = glmnet(x2m4, ym4, alpha=1, lambda = part_best.lasso.lambdam4)
part_lasso.ypredm4 = predict(part_lasso_modelm4, newx = x2m4, lambda = part_best.lasso.lambdam4)
part_lasso.msem4 = mean((part_lasso.ypredm4 - ym4)^2)
paste0("Partial lasso RSME (-4): ", round(sqrt(part_lasso.msem4), 2))
```

\newpage
##Predictions of best model (RF)
```{r}
test_preds = predict(rf_model, test)
predictions = data.frame(id = test$ID, prediction = test_preds)
predictions = round(predictions,2)
write.csv(predictions, "testing_predictions_KWEE_ELIZABETH_ELK204.csv", row.names=FALSE)
head(predictions)
```
